Type theories as quotient inductive-recursive types

### 2025-01-21 tex/main.tex
Polish some paragraphs in S3.1

\subsection{Plan of the paper}
\subsection{Related work}
\paragraph*{Formalisation of type theory in type theory}
\cite{Danielsson2006,Altenkirch2016a}
\cite{Altenkirch2017}
\cite{Munoz1998}
\cite{Dybjer1996,Castellan2021}
\LT{Dybjer \cite{Dybjer1996} already says cwf is inductive-recursive}

@@

Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and a type under $\Delta$ to form a type under $\Gamma$, so type substitution as a constructor has the type
\begin{alignat*}{3}
  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma.
\end{alignat*}
If we were defining cwfs, it would be enough to require type substitution for $\Ty$.
However, we need a ground type $\UU$ in $\Ty$ for any context $\Gamma$ for our inductive definition (i.e.\ the initial cwf), 
otherwise the type $\Ty$ would be empty as well as other types:
\begin{alignat*}{3}
  \UU & : \implicit{\Gamma} & \Ty\, \Gamma.
\end{alignat*}

Note that $\Gamma$ and $\Delta$ (coloured {\color{gray}grey}) above are quantified implicitly.
For now, $\UU$ serves as a constant in $\Ty$, but we will reuse $\UU$ for the type of small types later.

@@

The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms, even for simple facts.
\begin{example}\label{ex:pi2-comp}
  Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $[\sigma;\tau] A$ under the context $\Gamma$, and this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ after applying the substitution $\sigma$. 
  In short, the following equality apparently holds
  \[
    \pi_2\,(\sigma ; \tau) = [\sigma] (\pi_2\,\tau)
  \]
  by a back-of-the-envelope calculation
  \begin{equation} \label{eq:pi2-comp-proof}
    \pi_2\,(\sigma ; \tau) 
    = \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))
    = \pi_2\,(\sigma;\pi_1\,\tau, [\sigma]\,(\pi_2\,\tau))
    = [\sigma] (\pi_2\,\tau).
  \end{equation}
  Yet, the left-hand side is a term of type $[\pi_1\,(\sigma;\tau)] A$, but the other is $[\sigma] [\pi_1\,\tau] A$.
  Hence the above identity does not even make sense, since their types do not match.
  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} [\sigma] (\pi_2\,\tau)$ or, equivalently
  \begin{equation}\label{eq:pi2-comp-real-proof}
    ([\pi_1(\sigma; \tau)]\,A, \pi_2(\sigma; \tau)) =^{(A : \Ty\,\Gamma) \times (\Tm \Gamma A)} ([\sigma]\,[\pi_1\,\tau]A, [\sigma] (\pi_2\,\tau))
  \end{equation}
  as inhabitants of a $\Sigma$-type, so we can reason about term equalities along with type equalities.
  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to eliminate that $\transp$ to derive the right hand side.

  To better illustrate the annoyance, note that a complete proof of \eqref{eq:pi2-comp-real-proof} requires us to show each of following equations:
  \begin{alignat*}{3}
         & ([\pi_1(\sigma; \tau)]\,A                    &&, \pi_2\,(\sigma ; \tau)) \\
    = {} & ([\pi_1(\sigma; (\pi_1\tau , \pi_2\tau))]\,A &&, \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))) \\
    = {} & ([\pi_1(\sigma;\pi_1\,\tau, \transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, \transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \\
    = {} & ([\sigma;\pi_1\,\tau]\,A &&, \transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \nonumber \\
    = {} & (\sub{\sigma}{\sub{\pi_1\tau}{A}} &&, [\sigma] (\pi_2\,\tau)). \nonumber
  \end{alignat*}
  The first three equations follow from Lemma~2.3.4in \cite{UFP2013}, while the last equation has nothing to do with substitution calculus but a property of transport (\cite[Lemma~2.3.2]{UFP2013}).
\end{example}

@@

One possibility is to define substitution calculus as a quotient inductive-inductive-\emph{recursive} type whereas type substitution as a simultaneous recursive function should compute.
Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ along do not make $[\blank]_{\Ty}\blank$ a function, we may define it by
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU & = \UU
\end{alignat*}
However, this definition is \emph{not} inductive-recursive in the style of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}: in the schema of Dybjer and Setzer, the codomain of a simultaneous function does not refer to any inductive type being defined.
While its general schema remains underdeveloped, this definition is accepted in proof assistants such as \Agda and used in other formalisation of type theory~\cite{Danielsson2006,Chapman2009}.
Still, although $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ can be proved equal to $A$ and $[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, they are not definitional but propositional.
This definition fails to eliminate any transport in the definition of substitution calculus.

@@

Yet, as we are defining a function on an inductive type with equality constructors, we have to prove that $[\blank]_{\Ty}\blank$ is coherent.\LT{general condition?} 
This boils down to that $[\blank]_{\Ty}\blank$ preserves equality constructors for quotient types.
\begin{proposition}[Coherence]
  $[\sigma]_{\Ty}\,A$ is (propositionally) equal to $[\tau]_{\Ty}\,A$ whenever $\sigma = \tau$.
  \danger
\end{proposition}
\begin{proof}
  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become definitional, it suffices to show that following identities
  \begin{align*}
    & [\sigma;\idS]_{\Ty} \;A = [\sigma]_{\Ty}\;A,
    && [\idS;\sigma]_{\Ty} \;A = [\sigma]_{\Ty}\;A, && [(\sigma;\tau);\gamma]_{\Ty} \;A = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
    & [\pi_1(\sigma, t)]_{\Ty}\;A = [\sigma]_{\Ty}\;A \\
    & [\sigma; (\tau, t)]_{\Ty} \;A = [(\sigma;\tau), t]_{\Ty}\;A && [\sigma]_{\Ty} \;A = [\emptyctx]_{\Ty}\;A, && [\sigma]_{\Ty} \;A = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
  \end{align*}
  hold (propositionally).
  The first two identities of the first row hold definitionally by \eqref{def:definitional-type-subst-1} and \eqref{def:definitional-type-subst-2}, while terms of the both sides of the third compute to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{def:definitional-type-subst-2}.

@@

  We show the remaining cases by induction on $A$.
  However, $\UU$ is the only constructor for $\Ty\;\Gamma$, and each of them computes to $\UU$ definitionally.
\end{proof}

@@

\section{Future work}

\paragraph*{Type theory in (observational) type theory}

\paragraph*{A general schema of quotient inductive-inductive-recursive types}
\subparagraph*{Formalising definitions by overlapping patterns}

@@

\section{Formal definitions in \Agda}
\end{document}

>>>>

\subsection{Plan of the paper}

@@

Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and a type under $\Delta$ to form a type under $\Gamma$, so type substitution as a constructor has the type
\begin{alignat*}{3}
  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma.
\end{alignat*}
where $\Gamma$ and $\Delta$ (coloured {\color{gray}grey}) above are quantified implicitly.

If we were defining cwfs, type substitution would be enough for $\Ty$.
However, we need a base case in $\Ty$ for any context $\Gamma$ for our inductive definition (i.e.\ the initial cwf), 
otherwise the type $\Ty$ would be empty as well as other types.
Hence another constructor is introduced:
\begin{alignat*}{3}
  \UU & : \implicit{\Gamma} & \Ty\, \Gamma.
\end{alignat*}
For now, $\UU$ serves as a constant in $\Ty$, but we will reuse $\UU$ for the type of small types later.

@@

The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms even for such a simple fact below.
\begin{example}\label{ex:pi2-comp}
  Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $[\sigma;\tau] A$ under the context $\Gamma$, and this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ after applying the substitution $\sigma$. 
  In short, the following equality apparently holds
  \[
    \pi_2\,(\sigma ; \tau) = [\sigma] (\pi_2\,\tau)
  \]
  by a back-of-the-envelope calculation
  \begin{equation} \label{eq:pi2-comp-proof}
    \pi_2\,(\sigma ; \tau) 
    = \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))
    = \pi_2\,(\sigma;\pi_1\,\tau, [\sigma]\,(\pi_2\,\tau))
    = [\sigma] (\pi_2\,\tau).
  \end{equation}
  Yet, the left-hand side is a term of type $[\pi_1\,(\sigma;\tau)] A$, but the other is $[\sigma] [\pi_1\,\tau] A$.
  Hence the above identity does not even make sense, since their types do not match.
  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} [\sigma] (\pi_2\,\tau)$ or, equivalently
  \begin{equation}\label{eq:pi2-comp-real-proof}
    ([\pi_1(\sigma; \tau)]\,A, \pi_2(\sigma; \tau)) =^{(A : \Ty\,\Gamma) \times (\Tm \Gamma A)} ([\sigma]\,[\pi_1\,\tau]A, [\sigma] (\pi_2\,\tau))
  \end{equation}
  as inhabitants of a $\Sigma$-type, so we can reason about term equalities along with type equalities.
  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to eliminate that $\transp$ to derive the right hand side.

  To better illustrate the annoyance, note that a complete proof of \eqref{eq:pi2-comp-real-proof} requires us to show each of following equations:
  \begin{alignat*}{3}
         & ([\pi_1(\sigma; \tau)]\,A                    &&, \pi_2\,(\sigma ; \tau)) \\
    = {} & ([\pi_1(\sigma; (\pi_1\tau , \pi_2\tau))]\,A &&, \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))) \\
    = {} & ([\pi_1(\sigma;\pi_1\,\tau, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \\
    = {} & ([\sigma;\pi_1\,\tau]\,A &&, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \nonumber \\
    = {} & (\sub{\sigma}{\sub{\pi_1\tau}{A}} &&, [\sigma] (\pi_2\,\tau)). \nonumber
  \end{alignat*}
  The first three equations follow from Lemma~2.3.4 in \cite{UFP2013}, while the last equation has nothing to do with substitution calculus but a property of transport (\cite[Lemma~2.3.2]{UFP2013}).
\end{example}

@@

One possibility is to define substitution calculus as a quotient inductive-inductive-\emph{recursive} type whereas type substitution as a simultaneous recursive function should compute.
Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ along do not make $[\blank]_{\Ty}\blank$ a function, we may define it by
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU & = \UU
\end{alignat*}
However, this definition is \emph{not} inductive-recursive in the sense of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}: in their schema, the codomain of a simultaneous recursive function does not refer to any inductive type being defined.
While its general schema remains underdeveloped, this definition is accepted in proof assistants such as \Agda and used in other formalisation of type theory~\cite{Danielsson2006}.
Still, although $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ can be proved equal to $A$ and $[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, they are not definitional but propositional.
This definition fails to eliminate any transport in the definition of substitution calculus.

@@

Yet, as we are defining a function on an inductive type with equality constructors, we have to prove that $[\blank]_{\Ty}\blank$ is coherent.\LT{general condition?} 
This boils down to that $[\blank]_{\Ty}\blank$ preserves equality constructors for quotient types.
\begin{proposition}[Coherence]
  $[\sigma]_{\Ty}\,A$ is (propositionally) equal to $[\tau]_{\Ty}\,A$ whenever $\sigma = \tau$.
  \danger
\end{proposition}
\begin{proof}
  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become definitional, it suffices to show that following identities
  \begin{align*}
    & [\sigma;\idS]_{\Ty} \;A = [\sigma]_{\Ty}\;A,
    && [\idS;\sigma]_{\Ty} \;A = [\sigma]_{\Ty}\;A, && [(\sigma;\tau);\gamma]_{\Ty} \;A = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
    & [\pi_1(\sigma, t)]_{\Ty}\;A = [\sigma]_{\Ty}\;A \\
    & [\sigma; (\tau, t)]_{\Ty} \;A = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A && [\sigma]_{\Ty} \;A = [\emptyctx]_{\Ty}\;A, && [\sigma]_{\Ty} \;A = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
  \end{align*}
  hold (propositionally).
  The first two identities of the first row hold definitionally by \eqref{def:definitional-type-subst-1} and \eqref{def:definitional-type-subst-2}, while terms of the both sides of the third compute to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{def:definitional-type-subst-2}.

@@

  We show the remaining cases by induction on $A$.
  However, $\UU$ is the only constructor for $\Ty\;\Gamma$ for any $\Gamma$ and each of identities with $A \defeq \UU$ computes to $\UU$ definitionally.
\end{proof}

@@

\section{Discussion and future work}
\subsection{Related work}
\paragraph*{Formalisation of type theory in type theory}
\cite{Danielsson2006,Altenkirch2016a,Chapman2009}
\cite{Altenkirch2017}
\cite{Munoz1998}
\cite{Dybjer1996,Castellan2021}
\LT{Dybjer \cite{Dybjer1996} already says cwf is inductive-recursive}

\subsection{Future work}
\paragraph*{Type theory in (observational) type theory}
\paragraph*{Formalising definitions by overlapping patterns}
\paragraph*{A general schema of quotient inductive-inductive-recursive types}

@@

\section{Formal definitions in Agda}
\end{document}
